---
title: "Untitled"
output: html_document
---

# Load packages

Run the following code in your console to install packages:

```
if (!require(htmlwidgets)) install.packages("htmlwidgets")
if (!require(wordcloud2)) install.packages("wordcloud2")
if (!require(GGally)) install.packages("GGally")
if (!require(plotly)) install.packages("plotly")
if (!require(ggmap)) install.packages("ggmap")
if (!require(maps)) install.packages("maps")
if (!require(networkD3)) install.packages("networkD3")
if (!require(webshot)) install.packages("webshot")
```

Load packages:

```{r}
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(gapminder))
suppressPackageStartupMessages(library(wordcloud2))
suppressPackageStartupMessages(library(GGally))
suppressPackageStartupMessages(library(scales))
suppressPackageStartupMessages(library(plotly))
suppressPackageStartupMessages(library(ggmap))
suppressPackageStartupMessages(library(maps))
suppressPackageStartupMessages(library(networkD3))
```

# Specialized Plot Types

## Maps

### Raster maps with `ggmap`

Steps to using `ggmap`:

1.  Load a map with `get_map`.
    - Special cases: `get_googlemap` (satellite? road? your choice.), `get_cloudmademap`, `get_stamenmap`
    - Google maps now requires API key! See `ggmap` github page for details.
2.  Feed that into the `ggmap` function to display the map.
    -   To make displays overtop of the map, add your `ggplot` call to the `base_layer` argument of `ggmap`, then add your layers as usual *outside* of the `ggmap` call.

Example:

```{r}
# crime data from `ggmap` package.
small_crime <- sample_n(crime, 10000)
bbox <- c(
    left   = -96, 
    bottom = 29.5, 
    right  = -95, 
    top    = 30
)
(basemap <- ggmap::get_stamenmap(bbox, maptype = "toner-lite", zoom = 9) %>% 
        ggmap::ggmap())
basemap + 
    geom_density2d(aes(lon, lat), 
                   data   = small_crime, 
                   colour = "red")
```

Exercise: Modify the following code to retrieve and plot map of UBC. Lower-left GPS coordinates (latitude, longitude): 49.241284, -123.273007; upper-right GPS coordinates: 49.281994, -123.216366. __WARNING__: I don't recommend trying zoom > 14 (it'll be slow).

```{r}
c(-123.273007, 49.241284, FILL_THIS_IN, FILL_THIS_IN) %>% 
    get_stamenmap(maptype = "toner-lite", FILL_THIS_IN) %>% 
    FILL_THIS_IN
```


### Vector maps

Maps are depicted as polygons. The `maps` package has lots of polygon data stored, and can be plotted with `maps::map()`

```{r}
maps::map("world")
maps::map("world", region="canada")
maps::map("france")
```

Get the data in tidy tibble format with `ggplot2::map_data()`, ready to use with `ggplot()`. Example:

```{r}
(world_dat <- ggplot2::map_data("world") %>% 
     as_tibble())
```

Notice the order of the points. Notice the groups. Use `geom_polygon`:

```{r}
ggplot(world_dat, aes(long, lat)) +
    geom_polygon(aes(group=group))
```

Exercise: Make a plot of Italy.

```{r}
FILL_THIS_IN
```


## Plotly

The `plotly` R package makes highlight-interactivity possible.

Consider the following plot:

```{r}
(p <- gapminder %>% 
     filter(continent != "Oceania") %>% 
     ggplot(aes(gdpPercap, lifeExp)) +
     geom_point(aes(colour=pop), alpha=0.2) +
     scale_x_log10(labels=dollar_format()) +
     scale_colour_distiller(
         trans   = "log10",
         breaks  = 10^(1:10),
         labels  = comma_format(),
         palette = "Greens"
     ) +
     facet_wrap(~ continent) +
     scale_y_continuous(breaks=10*(1:10)) +
     theme_bw())
```

1. Convert it to a `plotly` object by applying the `ggplotly()` function:

```{r}
FILL_THIS_IN
```

2. You can save a plotly graph locally as an html file. Try saving the above:
    - NOTE: plotly graphs don't seem to show up in Rmd _notebooks_, but they do with Rmd _documents_.

```{r}
p %>% 
    ggplotly() %>% 
    htmlwidgets::saveWidget("LOCATION_GOES_HERE")
```


3. Run this code to see the json format underneath:

```{r, results='hide'}
p %>% 
    ggplotly() %>% 
    plotly_json()
```


4. Check out code to make a plotly object from scratch using `plot_ly()` -- scatterplot of gdpPercap vs lifeExp.
    - Check out the [cheat sheet](https://images.plot.ly/plotly-documentation/images/r_cheat_sheet.pdf).

```{r}
plot_ly(gapminder, 
        x = ~gdpPercap, 
        y = ~lifeExp, 
        type = "scatter",
        mode = "markers",
        opacity = 0.2) %>% 
    layout(xaxis = list(type = "log"))
```

5. Add population to form a z-axis for a 3D plot:

```{r}
plot_ly(gapminder,
        x = ~gdpPercap,
        y = ~lifeExp,
        z = FILL_THIS_IN,
        type = "scatter3d",
        mode = "markers",
        opacity = 0.2)
```


## Wordclouds

The `wordcloud2` package is excellent for creating word clouds (with some interactivity!).

Check out the [vignette](https://cran.r-project.org/web/packages/wordcloud2/vignettes/wordcloud.html) or the [`wordcloud2` README](https://github.com/Lchiffon/wordcloud2/blob/master/readme.md). Here's an example from the latter:

```{r, echo=FALSE}
demoFreq %>% 
    filter(freq > 1) %>%
    wordcloud2(size = 1, minRotation = -pi/2, maxRotation = -pi/2, shape="circle")
```

Notice:

- arrangement of words on the plane
- angle of words
- colour
- size

Exercise: What am I saying a lot of in this document (aside from stopwords)? The wrangling is done for you.

```{r}
this_file <- read_file("lec4.Rmd") %>% 
    strsplit(" ") %>% 
    `[[`(1) %>% 
    as_tibble() %>% 
    count(value) %>% 
    setNames(c("word", "freq")) %>% 
    anti_join(get_stopwords()) %>% 
    filter(!(word %in% c("", "%>%", "=", "+", "\n", "<-", "-", "+\n")))
FILL_THIS_IN
```


## GGally

Comes with many extensions to `ggplot2`. Most useful being the pairs functions.

It's name is derived from the package's intent to be an "ally" of `ggplot2`.

Example: use `ggpairs` to plot all combinations of measurements of flea body parts (subset to the first 4 columns to save time):

```{r}
GGally::ggpairs(flea[, 1:4])
```

Or, use `ggscatmat` for numeric-only data. It's faster. Numeric gapminder data:

```{r}
gapminder %>% 
    select(lifeExp, gdpPercap, pop) %>% 
    GGally::ggscatmat()
```

Check out all dependencies in the `iris` data set:

```{r}
FILL_THIS_IN
```


## Network Diagrams

There are plenty of options for making network diagrams in R. A comprehensive overview of the landscape is given at [kateto.net](http://kateto.net/network-visualization).

We'll be looking at:

- `igraph`: a flexible package for handling network data. [website](http://igraph.org/); [tutorial](https://rstudio-pubs-static.s3.amazonaws.com/74248_3bd99f966ed94a91b36d39d8f21e3dc3.html)
- `networkD3`: a D3-backed package allowing interactivity. [tutorial](http://christophergandrud.github.io/networkD3/)

### Network Data

Example: Twitter. Let's make a simple example on the board!

Network data is comprised of:

- Nodes/vertices
- Edges/links

Storage of these data:

- Node data: variables describing each node.
    - Typical tidy data.
- Edge data:
    - Each edge indicated by two columns of nodes; other columns optionally include edge properties (like length).
    - Could also be in sparse/adjacency matrix format ("wide", or "untidy" data).

Example of Les Mis characters:

```{r}
data("MisLinks")
data("MisNodes")
(MisLinks <- as_tibble(MisLinks))
(MisNodes <- as_tibble(MisNodes))
```

### Generic Graphing

Make a simple interactive network plot out of the edges:

```{r}
networkD3::simpleNetwork(MisLinks)
```

Get more sophisticated with `forceNetwork` (group specification is required):

```{r}
networkD3::forceNetwork(MisLinks, MisNodes, NodeID="name", Group="group")
```

Make a graph with `igraph` of the example we did on the board:

```{r}
edges <- FILL_IN_WITH_SOME_EDGE_VECTOR
(g <- igraph::graph(edges))
plot(g)
```

### Specialized Network Diagrams

Check out this [tutorial starting at "sankey"](http://christophergandrud.github.io/networkD3/#sankey) for other types of specialized plots.

# If time remains: other `ggplot2` goodies (from last class)

## Saving images

- `knit`ing to html: figures are built-in.
- `knit`ing to md: figures are saved to file, and read in.
- Save to file from code:
    - `ggsave` for ggplot2 objects; 
    - `pdf()`/`png()`/... + `dev.off()` is the Base R way. 

You can also save an html page snapshot to png -- useful for taking a snapshot of interactive images. Here's an example of saving a snapshot of google.ca: 

```{r}
#webshot::install_phantomjs()
webshot::webshot("https://www.google.ca/", "google.pdf")
```


## Scales; Colour

Scale functions in `ggplot2` take the form `scale_[aesthetic]_[mapping]()`.

Let's first focus on the following plot:

```{r}
p_scales <- ggplot(gapminder, aes(gdpPercap, lifeExp)) +
     geom_point(aes(colour=pop), alpha=0.2)
p_scales + 
    scale_x_log10(labels=scales::dollar_format()) +
    scale_colour_continuous(
        trans="log10",
        labels=scales::comma_format()
    )
```

1. Change the y-axis tick mark spacing to 10; the colour spacing has been changed to include all powers of 10.

```{r}
p_scales +
    scale_x_log10(labels=scales::dollar_format()) +
    scale_colour_continuous(
        trans  = "log10", 
        labels = scales::comma_format(),
        breaks = 10^(1:10)
    ) +
    FILL_IN_SCALE_FUNCTION(breaks=FILL_IN_BREAKS)
```


2. Use `RColorBrewer` to change the colour scheme.
    - Notice the three different types of scales: sequential, diverging, and continuous.

```{r}
## All palettes the come with RColorBrewer:
RColorBrewer::display.brewer.all()
p_scales +
    scale_x_log10(labels=scales::dollar_format()) +
    FILL_IN_WITH_RCOLORBREWER(
        trans   = "log10",
        breaks  = 10^(1:10),
        labels  = scales::comma_format(),
        palette = FILL_THIS_IN
    ) +
    scale_y_continuous(breaks=10*(1:10))
```

3. Run the following code to check out the `viridis` scale for a colour-blind friendly scheme.
    - Hint: add `scale_colour_viridis_c` (`c` stands for continuous; `d` discrete).
    - You can choose a palette with `option`.

```{r}
p_scales +
    scale_x_log10(labels=scales::dollar_format()) +
    scale_colour_viridis_c(
        trans   = "log10",
        breaks  = 10^(1:10),
        labels  = scales::comma_format()
    ) +
    scale_y_continuous(breaks=10*(1:10))
```


## Fill, or Colour?

For each (non-Oceania) continent split into high and low life expectancies, make a tile plot where colour represents mean GDP per capita.

- Try `colour`/`color` and `fill`. 

```{r}
gapminder %>% 
    filter(continent != "Oceania") %>% 
    mutate(qualLifeExp = if_else(lifeExp > 60, "high", "low")) %>% 
    group_by(continent, qualLifeExp) %>% 
    summarize(gdpPercap = mean(gdpPercap)) %>% 
    ggplot(aes(continent, qualLifeExp)) +
    geom_tile(FILL_THIS_IN)
```

On the plot of lifeExp vs gdpPercap, colour points by continent.

- Try `shape=21` to distinguish between `fill` (interior) and `colour` (exterior).

```{r}
ggplot(gapminder, aes(gdpPercap, lifeExp)) +
    scale_x_log10() +
    geom_point(FILL_THIS_IN)
```

List of shapes can be found [at the bottom of the `scale_shape` documentation](https://ggplot2.tidyverse.org/reference/scale_shape.html).


## Overplotted scatterplot

Try a scatterplot with:

- Alpha transparency.
- `geom_hex()`
- `geom_density2d()`

```{r}
ggplot(gapminder, aes(gdpPercap, lifeExp)) +
    ADD_GEOM_HERE
```


## Polar coordinates

Change the following coordinates to polar:

```{r}
gapminder %>% 
    filter(year == 2007, continent == "Africa") %>% 
    ggplot(aes(country, lifeExp)) + 
    geom_col()
```
