---
title: "Lecture 3 Worksheet"
output: html_document
---

```{r}
library(tidyverse)
library(gapminder)
library(RColorBrewer)
library(scales)
```


## Facetting

Make histograms of `gdpPercap` for each (non-Oceania) continent by adding a line to the following code. 

- Try facetting by `qualLifeExp`. 
- Try the `scales` and `ncol` arguments of the facet layer. 

```{r}
gapminder %>% 
    filter(continent != "Oceania") %>% 
    mutate(qualLifeExp = if_else(lifeExp > 60, "high", "low")) %>% 
    ggplot(aes(x=gdpPercap)) +
    geom_histogram() +
    scale_x_log10()
```

| Grammar Component     | Specification |
|-----------------------|---------------|
| __data__              | `gapminder` |
| statistical transform | histogram (binning and counting) |
| __aesthetic mapping__ | x=gdpPercap; y=count |
| __geometric object__  | histogram |
| scale                 | x is log10; y is linear. |
| coordinate system     | Rectangular/Cartesian |
| facetting             | FILL_THIS_IN |

# Theming

__Question__: What makes a plot "publication quality"?

Changing the look of a graphic can be achieved through the `theme()` layer.

There are ["complete themes"](http://ggplot2.tidyverse.org/reference/ggtheme.html) that come with `ggplot2`, my favourite being `theme_bw` (I've grown tired of the default gray background, so `theme_bw` is refreshing).

1. Change the theme of the following plot to `theme_bw()`:

```{r}
ggplot(iris, aes(Sepal.Width, Sepal.Length)) +
     facet_wrap(~ Species) +
     geom_point() +
     labs(x = "Sepal Width",
          y = "Sepal Length",
          title = "Sepal sizes of three plant species")
```

2. Then, change font size of axis labels, and the strip background colour. Others?

```{r}
ggplot(iris, aes(Sepal.Width, Sepal.Length)) +
     facet_wrap(~ Species) +
     geom_point() +
     labs(x = "Sepal Width",
          y = "Sepal Length",
          title = "Sepal sizes of three plant species") +
    theme_bw() +
    theme(FILL_IN_ARG_FOR_AXIS_TITLE = FILL_THIS_IN,
          strip.background = FILL_THIS_IN)
```

# Scales; Colour

Scale functions in `ggplot2` take the form `scale_[aesthetic]_[mapping]()`.

Let's first focus on the following plot:

```{r}
p_scales <- ggplot(gapminder, aes(gdpPercap, lifeExp)) +
     geom_point(aes(colour=pop), alpha=0.2)
p_scales + 
    scale_x_log10() +
    scale_colour_continuous(trans="log10")
```

1. Change the y-axis tick mark spacing to 10; change the colour spacing to include all powers of 10.

```{r}
p_scales +
    scale_x_log10() +
    scale_colour_continuous(
        trans  = "log10", 
        breaks = FILL_IN_BREAKS
    ) +
    FILL_IN_SCALE_FUNCTION(breaks=FILL_IN_BREAKS)
```

2. Specify `scales::*_format` in the `labels` argument of a scale function to do the following:
    - Change the x-axis labels to dollar format (use `scales::dollar_format()`)
    - Change the colour labels to comma format (use `scales::comma_format()`)

```{r}
p_scales +
    scale_x_log10(labels=FILL_THIS_IN) +
    scale_colour_continuous(
        trans  = "log10", 
        breaks = 10^(1:10),
        labels = FILL_THIS_IN
    ) +
    scale_y_continuous(breaks=10*(1:10))
```

3. Use `RColorBrewer` to change the colour scheme.
    - Notice the three different types of scales: sequential, diverging, and continuous.

```{r}
## All palettes the come with RColorBrewer:
RColorBrewer::display.brewer.all()
p_scales +
    scale_x_log10(labels=dollar_format()) +
    FILL_IN_WITH_RCOLORBREWER(
        trans   = "log10",
        breaks  = 10^(1:10),
        labels  = comma_format(),
        palette = FILL_THIS_IN
    ) +
    scale_y_continuous(breaks=10*(1:10))
```

4. Run the following code to check out the `viridis` scale for a colour-blind friendly scheme.
    - Hint: add `scale_colour_viridis_c` (`c` stands for continuous; `d` discrete).
    - You can choose a palette with `option`.

```{r}
p_scales +
    scale_x_log10(labels=dollar_format()) +
    scale_colour_viridis_c(
        trans   = "log10",
        breaks  = 10^(1:10),
        labels  = comma_format()
    ) +
    scale_y_continuous(breaks=10*(1:10))
```

## Requests?

Is there anything you'd like to see done in `ggplot2`? Keep in mind we're seeing special plot types next time.


## Bubble Plots

Have three numeric variables? Try mapping to size.

Example: x=`gdpPercap`, y=`lifeExp`, size/cex=`pop`

- Try adding `scale_radius()` and `scale_size_area()`. What's better?

```{r}
ggplot(gapminder, aes(gdpPercap, lifeExp)) +
    scale_x_log10() +
    geom_point(aes(FILL_THIS_IN))
```

Try now making `pop` a colour/color:

```{r}
ggplot(gapminder, aes(gdpPercap, lifeExp)) +
    scale_x_log10() +
    geom_point(aes(FILL_THIS_IN)) +
    #scale_colour_continuous(trans="log10")
```


## Fill, or Colour?

For each (non-Oceania) continent split into high and low life expectancies, make a tile plot where colour represents mean GDP per capita.

- Try `colour`/`color` and `fill`. 

```{r}
gapminder %>% 
    filter(continent != "Oceania") %>% 
    mutate(qualLifeExp = if_else(lifeExp > 60, "high", "low")) %>% 
    group_by(continent, qualLifeExp) %>% 
    summarize(gdpPercap = mean(gdpPercap)) %>% 
    ggplot(aes(continent, qualLifeExp)) +
    geom_tile(FILL_THIS_IN)
```

On the plot of lifeExp vs gdpPercap, colour points by continent.

- Try `shape=21` to distinguish between `fill` (interior) and `colour` (exterior).

```{r}
ggplot(gapminder, aes(gdpPercap, lifeExp)) +
    scale_x_log10() +
    geom_point(FILL_THIS_IN)
```

List of shapes can be found [at the bottom of the `scale_shape` documentation](https://ggplot2.tidyverse.org/reference/scale_shape.html).


## Overplotted scatterplot

Try a scatterplot with:

- Alpha transparency.
- `geom_hex()`
- `geom_density2d()`

```{r}
ggplot(gapminder, aes(gdpPercap, lifeExp)) +
    ADD_GEOM_HERE
```



## Polar coordinates

Change the following coordinates to polar:

```{r}
gapminder %>% 
    filter(year == 2007, continent == "Africa") %>% 
    ggplot(aes(country, lifeExp)) + 
    geom_col()
```

# Saving Graphs to File

- Don't use the mouse
- Use `ggsave` for ggplot
    - Practice by saving the following plot to file: 

```{r}
ggplot(mtcars, aes(hp, wt)) + 
    geom_point()
ggsave(FILENAME_HERE, PLOT_OBJECT_HERE)
```

- Base R way: print plots "to screen", sandwiched between `pdf()`/`jpeg()`/`png()`... and `dev.off()`. 
- Vector vs. raster: Images are stored on your computer as either _vector_ or _raster_.
    - __Raster__: an `n` by `m` grid of pixels, each with its own colour. `jpeg`, `png`, `gif`, `bmp`.
    - __Vector__: represented as shapes and lines. `pdf`, [`svg`](https://www.w3schools.com/graphics/svg_intro.asp).
    - For tips: ["10 tips for making your R graphics look their best""](http://blog.revolutionanalytics.com/2009/01/10-tips-for-making-your-r-graphics-look-their-best.html).
    





## Practice Exercises

__Exercise 1__: Make the following `gapminder` plot:

- scatterplot of `gdpPercap` vs `lifeExp`
- `pop` indicated by bubble area
- facet by `year`
- colour by `continent`

```{r}

```


__Exercise 2__: Make a plot of `year` (x) vs `lifeExp` (y), with points coloured by continent. Then, to that same plot, fit a straight regression line to each continent, without the error bars. If you can, try piping the data frame into the `ggplot` function.

```{r}

```


__Exercise 3__: Repeat Exercise 2, but switch the _regression line_ and _geom\_point_ layers. How is this plot different from that of Exercise 1?

```{r}

```


__Exercise 4__: Omit the `geom_point` layer from either of the above two plots (it doesn't matter which). Does the line still show up, even though the data aren't shown? Why or why not?

```{r}

```


__Exercise 5__: Make a plot of `year` (x) vs `lifeExp` (y), facetted by continent. Then, fit a smoother through the data for each continent, without the error bars. Choose a span that you feel is appropriate.

```{r}

```


__Exercise 6__: Plot the population over time (year) using lines, so that each country has its own line. Colour by `gdpPercap`. Add alpha transparency to your liking. 

```{r}

```


__Exercise 7__: Add points to the plot in Exercise 6.

```{r}

```

