---
title: "Lecture 1 Worksheet"
output: github_document
---

In this worksheet, we'll be exploring various plot types (i.e., geometric objects), only using the `x` and `y` aesthetics (and `group`).

We'll be jumping straight into the `ggplot()` function, instead of the more limited `qplot()` function.

1. To get started, load the `tidyverse` and `gapminder` R packages. 

```{r}
suppressPackageStartupMessages(library(tidyverse))
library(gapminder)
```


## Scatterplot

Let's look at a _scatterplot_ of `gdpPercap` vs. `lifeExp`. 

1. Fill out the grammar components below. Again, bold _must_ be specified to make a `ggplot2` plot.
    - We'll ignore "coordinate system" and "facetting" after this.

| Grammar Component     | Specification |
|-----------------------|---------------|
| __data__              | `gapminder` |
| __aesthetic mapping__ | x=lifeExp and y=gdpPercap  |
| __geometric object__  | point |
| scale                 | linear |
| statistical transform | none |
| coordinate system     | rectangular/cartesian |
| facetting             | none |

2. Populate the data and aesthetic mapping in `ggplot`. What is returned? What's missing?

```{r}
ggplot(data=gapminder, mapping=aes(x=lifeExp, y=gdpPercap)) +
    geom_point()
ggplot(gapminder, aes(lifeExp, gdpPercap)) +
    geom_point()
ggplot(gapminder) +
    geom_point(aes(x=lifeExp, y=gdpPercap))
```


~~3. Add the missing component as a _layer_.~~



Notice the "metaprogramming" again!

4. You _must_ remember to put the aesthetic mappings in the `aes` function! What happens if you forget?

```{r}
#ggplot(gapminder) +
#    geom_point(x = lifeExp, y = gdpPercap)
```


5. Put the x-axis on a log scale, first by transforming the x variable. 
    - Note: `ggplot2` does some data wrangling and computations itself! We don't always have to modify the data frame.

```{r}
ggplot(gapminder, aes(lifeExp, gdpPercap)) +
    geom_point() +
    scale_y_log10()

ggplot(gapminder, aes(lifeExp, log(gdpPercap))) +
    geom_point()
```


~~6. Try again, this time by changing the _scale_ (this way is better).~~



~~7. The aesthetic mappings can be specified on the geom layer if you want, instead of the main `ggplot` call. Give it a try:~~


8. Optional: git stage and commit

__Uses of a scatterplot__: 

- Visualize 2-dimensional distributions; dependence.
- 2 numeric variables

## Histograms, and Kernel Density Plots

Let's build a histogram of life expectancy.

1. Fill out the grammar components below. Again, bold _must_ be specified to make a `ggplot2` plot.

| Grammar Component     | Specification |
|-----------------------|---------------|
| __data__              | `gapminder` |
| __aesthetic mapping__ | x=lifeExp, y=count (__corrected from before__) |
| __geometric object__  | histogram |
| scale                 | x and y both linear |
| statistical transform | count |

2. Build the histogram of life expectancy.

```{r}
ggplot(gapminder, aes(lifeExp)) +
    geom_histogram(bins=50)
```


~~3. Change the number of bins to 50.~~



4. Instead of a histogram, let's create a kernel density plot.

```{r}
ggplot(gapminder, aes(lifeExp)) +
    geom_density()
```


5. Optional: git stage and commit

__Uses of a histogram__: Explore the distribution of a single numeric variable.


## Box plots, and violin plots

Let's make _box plots_ of population for each continent. Note: y-axis is much better on a log scale!

1. Fill out the grammar components below. Again, bold _must_ be specified to make a `ggplot2` plot.

| Grammar Component     | Specification |
|-----------------------|---------------|
| __data__              | `gapminder` |
| __aesthetic mapping__ | x=continent, y=gdpPercap |
| __geometric object__  | boxplot OR violin |
| scale                 | log-y; x is linear |
| statistical transform | boxplot: 5 number summary; violinplot: density estimate |

2. Initiate the `ggplot` call, with the log y scale, and store it in the variable `a`. Print out `a`.

```{r}
a <- ggplot(gapminder, aes(continent, gdpPercap)) +
    scale_y_log10()

```



3. Add the boxplot geom to `a`.

```{r}
a + geom_boxplot()
a + geom_point(alpha=0.2)
```


4. A violin plot is a kernel density on its side, made symmetric. Add that geom to `a`.
    - What's better here, boxplots or violin plots? Why?

```{r}
a + geom_violin()
ggplot(gapminder, aes(continent, lifeExp)) +
    geom_violin()
```


5. Optional: git stage and commit

__Use of boxplot__: Visualize 1-dimensional distributions (of a single numeric variable).

## Jitter plots

Let's work up to the concept of a _jitter plot_. As above, let's explore the population for each continent, but using points (again, with the y-axis on a log scale).

Let's hold off on identifying the grammar. 

1. Initiate the `ggplot` call to make a scatterplot of `continent` vs `pop`; initiate the log y scale. Store the call in the variable `b`.

```{r}
b <- ggplot(gapminder, aes(continent, pop)) +
    scale_y_log10()
```


2. Add the point geom to `b`. Why is this an ineffective plot?

```{r}
b + geom_point()
```


3. A solution is to jitter the points. Add the jitter geom. Re-run the command a few times -- does the plot change? Why?

```{r}
b + geom_jitter()
b + geom_violin() + geom_jitter(alpha=0.1)
```


4. How does the grammar differ from a box plot or violin plot?
    - ANSWER: 

5. We can add multiple geom _layers_ to our plot. Put a jitterplot overtop of the violin plot, starting with our base `b`. Try vice-versa. 



6. Optional: git stage and commit

__Uses of jitterplot__: Visualize 1-dimensional distributions, AND get a sense of the sample size.
