---
title: "Lecture 2 Worksheet"
output: github_document
---

```{r}
suppressPackageStartupMessages(library(tidyverse))
library(gapminder)
students <- as_tibble(HairEyeColor) %>% 
    uncount(n)
```

Note: `students` contains 592 observations of Hair, Eye, and Sex:

```{r}
students %>% 
    sample_n(10)
```

# Finish 1- and 2- variable plots from last time

## Bar plots

__Uses of bar plots__: 

1. Estimate probability mass functions / view frequencies of categories
    - One categorical variable.
2. Compare a single numeric response corresponding to different categories.
    - One categorical, one (unique) numeric variable.

What is the distribution of eye colour in `students`?

```{r}
ggplot(students, aes(Eye)) +
    geom_bar()
```

| Grammar Component     | Specification |
|-----------------------|---------------|
| __data__              | `students` |
| statistical transform | count |
| __aesthetic mapping__ | x=Eye; y=count |
| __geometric object__  | Bars |
| scale                 | Linear count |
| coordinate system     | Rectangular/Cartesian |
| facetting             | None |

How does the life expectancy of Canada, USA, and Mexico compare in 2007?

```{r}
(small_gap <- gapminder %>% 
     filter(country %in% c("Canada", "United States", "Mexico"),
            year == 2007) %>% 
     select(country, lifeExp))
ggplot(small_gap, aes(x=country, y=lifeExp)) +
    geom_col()
```

| Grammar Component     | Specification |
|-----------------------|---------------|
| __data__              | `students` |
| statistical transform | none |
| __aesthetic mapping__ | x=country, y=lifeExp |
| __geometric object__  | Bars |
| scale                 | Linear |
| coordinate system     | Rectangular/Cartesian |
| facetting             | None |

## Heatmaps

__Use of heatmaps__: show dependence amongst two categorical variables.

Example: Dependence amongst hair colour and eye colour in the `students` data.

- Points? Jitter? No.
- `geom_count()`?
- `geom_bin2d()`!

```{r}
heat <- ggplot(students, aes(Hair, Eye))
heat + geom_jitter()
heat + geom_count()
heat + geom_bin2d()
```

Fill in the grammar components:

| Grammar Component     | Specification |
|-----------------------|---------------|
| __data__              | `students` |
| statistical transform | count |
| __aesthetic mapping__ | x=Hair, y=Eye; colour=count |
| __geometric object__  | rectangles/squares |
| scale                 | Linear count |
| coordinate system     | Rectangular/Cartesian |
| facetting             | None |

# Three+ Variable Plots

## Time/Line Plots

__Uses of time/line plots__: 

- Visualize trends of a numeric variable over time.

Plot life expectancy over time for each country in `gapminder`.

```{r}
ggplot(gapminder, aes(year, lifeExp)) +
    geom_line(aes(group=country, colour=continent), alpha=0.2)
```

| Grammar Component     | Specification |
|-----------------------|---------------|
| __data__              | `gapminder` |
| statistical transform | none |
|__aesthetic mapping__|x=year, y=lifeExp, group=country, colour=continent |
| __geometric object__  | line |
| scale                 | x and y both linear |
| coordinate system     | Rectangular/Cartesian |
| facetting             | None |

Compare to __path plots__, for which the order matters:

Add to the following scatterplot to see how Rwanda's life expectancy and GDP per capita evolved over time:

```{r}
gapminder %>% 
    filter(country == "Rwanda") %>% 
    arrange(year) %>% 
    ggplot(aes(gdpPercap, lifeExp)) +
    geom_point() +
    geom_path(arrow=arrow())
```
