---
title: "Lecture 2 Worksheet"
output: github_document
---

```{r}
suppressPackageStartupMessages(library(tidyverse))
library(gapminder)
students <- as_tibble(HairEyeColor) %>% 
    uncount(n)
```

Note: `students` contains 592 observations of Hair, Eye, and Sex:

```{r}
students %>% 
    sample_n(10)
```

# Finish 1- and 2- variable plots from last time

## Bar plots

__Uses of bar plots__: 

1. Estimate probability mass functions / view frequencies of categories
    - One categorical variable.
2. Compare a single numeric response corresponding to different categories.
    - One categorical, one (unique) numeric variable.

What is the distribution of eye colour in `students`?

```{r}
ggplot(students, aes(FILL_THIS_IN)) +
    FILL_THIS_IN
```

| Grammar Component     | Specification |
|-----------------------|---------------|
| __data__              | `students` |
| statistical transform | FILL_THIS_IN |
| __aesthetic mapping__ | FILL_THIS_IN; TRANSFORM_VARIABLE_TOO |
| __geometric object__  | Bars |
| scale                 | Linear count |
| coordinate system     | Rectangular/Cartesian |
| facetting             | None |

How does the life expectancy of Canada, USA, and Mexico compare in 2007?

```{r}
(small_gap <- gapminder %>% 
     filter(country %in% c("Canada", "United States", "Mexico"),
            year == 2007) %>% 
     select(country, lifeExp))
ggplot(small_gap, aes(FILL_THIS_IN)) +
    FILL_THIS_IN
```

| Grammar Component     | Specification |
|-----------------------|---------------|
| __data__              | `students` |
| statistical transform | FILL_THIS_IN |
| __aesthetic mapping__ | FILL_THIS_IN |
| __geometric object__  | Bars |
| scale                 | Linear |
| coordinate system     | Rectangular/Cartesian |
| facetting             | None |

## Heatmaps

__Use of heatmaps__: show dependence amongst two categorical variables.

Example: Dependence amongst hair colour and eye colour in the `students` data.

- Points? Jitter? No.
- `geom_count()`?
- `geom_bin2d()`!

```{r}
heat <- ggplot(students, aes(Hair, Eye))
heat + FILL_THIS_IN
```

Fill in the grammar components:

| Grammar Component     | Specification |
|-----------------------|---------------|
| __data__              | `students` |
| statistical transform | FILL_THIS_IN |
| __aesthetic mapping__ | x=Hair, y=Eye; INDICATE_TRANSFORM_VARIABLE_HERE |
| __geometric object__  | rectangles/squares |
| scale                 | Linear count |
| coordinate system     | Rectangular/Cartesian |
| facetting             | None |

# Three+ Variable Plots

## Time/Line Plots

__Uses of time/line plots__: 

- Visualize trends of a numeric variable over time.

Plot life expectancy over time for each country in `gapminder`.

```{r}
ggplot(gapminder, aes(year, lifeExp)) +
    FILL_THIS_IN
```

| Grammar Component     | Specification |
|-----------------------|---------------|
| __data__              | `gapminder` |
| statistical transform | none |
| __aesthetic mapping__ | x=year, y=lifeExp, FILL_THIS_IN |
| __geometric object__  | FILL_THIS_IN |
| scale                 | x and y both linear |
| coordinate system     | Rectangular/Cartesian |
| facetting             | None |

Compare to __path plots__, for which the order matters:

Add to the following scatterplot to see how Rwanda's life expectancy and GDP per capita evolved over time:

```{r}
gapminder %>% 
    filter(country == "Rwanda") %>% 
    ggplot(aes(gdpPercap, lifeExp)) +
    geom_point()
```

## Facetting

Make histograms of `gdpPercap` for each (non-Oceania) continent by adding a line to the following code. 

- Try facetting by `qualLifeExp`. 
- Try the `scales` and `ncol` arguments of the facet layer. 

```{r}
gapminder %>% 
    filter(continent != "Oceania") %>% 
    mutate(qualLifeExp = if_else(lifeExp > 60, "high", "low")) %>% 
    ggplot(aes(x=gdpPercap)) +
    geom_histogram() +
    scale_x_log10()
```

| Grammar Component     | Specification |
|-----------------------|---------------|
| __data__              | `gapminder` |
| statistical transform | histogram (binning and counting) |
| __aesthetic mapping__ | x=gdpPercap |
| __geometric object__  | histogram |
| scale                 | x is log10; y is linear. |
| coordinate system     | Rectangular/Cartesian |
| facetting             | FILL_THIS_IN |


## Bubble Plots

Have three numeric variables? Try mapping to size.

Example: x=`gdpPercap`, y=`lifeExp`, size/cex=`pop`

- Try adding `scale_radius()` and `scale_size_area()`. What's better?

```{r}
ggplot(gapminder, aes(gdpPercap, lifeExp)) +
    scale_x_log10() +
    geom_point(aes(FILL_THIS_IN))
```

Try now making `pop` a colour/color:

```{r}
ggplot(gapminder, aes(gdpPercap, lifeExp)) +
    scale_x_log10() +
    geom_point(aes(FILL_THIS_IN)) +
    #scale_colour_continuous(trans="log10")
```


## Fill, or Colour?

For each (non-Oceania) continent split into high and low life expectancies, make a tile plot where colour represents mean GDP per capita.

- Try `colour`/`color` and `fill`. 

```{r}
gapminder %>% 
    filter(continent != "Oceania") %>% 
    mutate(qualLifeExp = if_else(lifeExp > 60, "high", "low")) %>% 
    group_by(continent, qualLifeExp) %>% 
    summarize(gdpPercap = mean(gdpPercap)) %>% 
    ggplot(aes(continent, qualLifeExp)) +
    geom_tile(FILL_THIS_IN)
```

On the plot of lifeExp vs gdpPercap, colour points by continent.

- Try `shape=21` to distinguish between `fill` (interior) and `colour` (exterior).

```{r}
ggplot(gapminder, aes(gdpPercap, lifeExp)) +
    scale_x_log10() +
    geom_point(FILL_THIS_IN)
```

List of shapes can be found [at the bottom of the `scale_shape` documentation](https://ggplot2.tidyverse.org/reference/scale_shape.html).

# Exercises

__Exercise 1__: Make the following `gapminder` plot:

- scatterplot of `gdpPercap` vs `lifeExp`
- `pop` indicated by bubble area
- facet by `year`
- colour by `continent`

```{r}

```


__Exercise 2__: Make a plot of `year` (x) vs `lifeExp` (y), with points coloured by continent. Then, to that same plot, fit a straight regression line to each continent, without the error bars. If you can, try piping the data frame into the `ggplot` function.

```{r}

```


__Exercise 3__: Repeat Exercise 1, but switch the _regression line_ and _geom\_point_ layers. How is this plot different from that of Exercise 1?

```{r}

```


__Exercise 4__: Omit the `geom_point` layer from either of the above two plots (it doesn't matter which). Does the line still show up, even though the data aren't shown? Why or why not?

```{r}

```


__Exercise 5__: Make a plot of `year` (x) vs `lifeExp` (y), facetted by continent. Then, fit a smoother through the data for each continent, without the error bars. Choose a span that you feel is appropriate.

```{r}

```


__Exercise 6__: Plot the population over time (year) using lines, so that each country has its own line. Colour by `gdpPercap`. Add alpha transparency to your liking. 

```{r}

```


__Exercise 7__: Add points to the plot in Exercise 5.

```{r}

```

